Reasoning Graphically about Quantum Computation

### 2008-02-28 papers/reasoning-graphically/paper.tex
more diagrams and some restructuring
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1388 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

\section{Quotients and Rewriting}
\label{sec:quotients-rewriting}

Few examples of interesting compact closed categories are freely
generated.  In practice we work with categories presented as
generators and \emph{equations}.  For 
representing quantum computations, Coecke and Duncan
\cite{Coecke:2008jo} propose the following collection of generators:
\begin{description}
\item[Objects] the self dual object $Q$;
\item[Arrows] Two families of arrows:
  \begin{align}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
We represent these two families, and their adjoints, by colours in our
graphical notation:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  The free
compact closed category  is then given by all graphs formed by
composing and tensoring these basic graphs.

To model the behaviour of quantum systems certain additional equations
must be satisfied.  Each family $(\delta, \epsilon)$ must form a
\emph{classical structure}
\cite{Coecke2006Quantum-Measure,Coecke2006POVMs-and-Naima}:  the pair
$(\delta, \epsilon)$  should form a cocommuative comonoid, the pair
$(\delta^\dag,\epsilon^\dag)$ must form a commutative monoid, and
together they must satisfy the isometry and frobenius equations. In
addition the family $\alpha$ must form an abelian group with $\alpha^\dag
= -\alpha$.  These equations are presented graphically below for the
green family;  the red vertices obey all the same equations
\begin{description}
\item[Comonoid Laws] 
\[
\begin{array}{ccccccccccccccc}
  \inlinegraphic{3em}{comonoid-assoc1}  &=\;&
  \inlinegraphic{3em}{comonoid-assoc2}
  &\qquad\qquad&
  \inlinegraphic{3em}{comonoid-unit1}  &=\;&
  \inlinegraphic{3em}{comonoid-unit2}  &=\;&
  \inlinegraphic{3em}{comonoid-unit3} 
  &\qquad\qquad&
  \inlinegraphic{3em}{comonoid-comm1}  &=\;&
  \inlinegraphic{3em}{comonoid-comm2}
\end{array}
\]
\item[Monoid Laws] 
\[
\begin{array}{ccccccccccccccc}
  \inlinegraphic{3em}{monoid-assoc1}  &=\;&
  \inlinegraphic{3em}{monoid-assoc2}
  &\qquad\qquad&
  \inlinegraphic{3em}{monoid-unit1}  &=\;&
  \inlinegraphic{3em}{monoid-unit2}  &=\;&
  \inlinegraphic{3em}{monoid-unit3} 
  &\qquad\qquad&
  \inlinegraphic{3em}{monoid-comm1}  &=\;&
  \inlinegraphic{3em}{monoid-comm2}
\end{array}
\]
\item[Isometry]
\[
\begin{array}{ccc}
  \inlinegraphic{3.5em}{isometry1}  &=\;&
  \inlinegraphic{3.5em}{isometry2}  
\end{array}
\]
\item[Frobenius]
  \[
  \begin{array}{ccccc}
    \inlinegraphic{3em}{frobenius1}  &=\;&
    \inlinegraphic{3em}{frobenius2} &=\;&
    \inlinegraphic{3em}{frobenius3}
  \end{array}
  \]
\item[Compact Structure] 
  \[
  \begin{array}{ccccccc}
    \inlinegraphic{2.5em}{compact1}  &=\;&
    \inlinegraphic{1.3em}{compact2}  
    &\qquad\qquad&
    \inlinegraphic{2.5em}{compact3}  &=\;&
    \inlinegraphic{1.3em}{compact4}  
  \end{array}
  \]
  \item[Abelian Unitary Group]
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{2.2em}{group1}  &:=\;&
      \inlinegraphic{2.2em}{group2}  &=\;&
      \inlinegraphic{2.2em}{group3} 
      &\qquad\qquad&
      \left( \inlinegraphic{2.2em}{group4}\!\right)^\dag  &=\;&
      \inlinegraphic{2.2em}{group5} 
      &\qquad\qquad&
      \inlinegraphic{3.5em}{group6}  &=\;&
      \inlinegraphic{3.5em}{group7}  &=\;&
      \inlinegraphic{3.5em}{group8} 
    \end{array}
    \]
  \item[Bilinearity] 
    \[
    \begin{array}{cccccccccccc}
      \inlinegraphic{3.5em}{alpha-commute1}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute2}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute3} 
      &\qquad\qquad&
      \inlinegraphic{3.5em}{alpha-commute4}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute5}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute6} 
    \end{array}
    \]
\end{description}
Additionally, the two families satisfy equations which make them a
\emph{scaled bialgebra}.
\begin{description}
\item[Bialgebra Laws] Let  $\inlinegraphic{1em}{bialgebra1} := 
      \inlinegraphic{1.5em}{bialgebra2}$;  then:
  \[
  \begin{array}{ccccccccccccccc}
      \inlinegraphic{3.5em}{bialgebra3}  &=\;&
      \inlinegraphic{3em}{bialgebra4}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra5}  &=\;&
      \inlinegraphic{1.5em}{bialgebra6}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra7}  &=\;&
      \inlinegraphic{1.5em}{bialgebra8} \\ \\
      \inlinegraphic{3.5em}{bialgebra9}  &=\;&
      \inlinegraphic{3em}{bialgebra10}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra11}  &=\;&
      \inlinegraphic{1.5em}{bialgebra12}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra13}  &=\;&
      \inlinegraphic{1.5em}{bialgebra14} 
  \end{array}
  \]
\end{description}
Last group of equations

Statement and Discussion of spider law.

The need to formalise these equations leads naturally to a rewriting
theory

>>>>

\section{Graphs as Quantum Computations}
\label{sec:quotients-rewriting}

While compact closed categories provide a suitable setting for
reasoning about quantum computation \cite{AbrCoe:CatSemQuant:2004},
freely generated structure will not suffice:  we need additional
equations.  These equations will be expressed as rewrites rules for
graphs.  In this section we will describe the generators and equations
used to reason about quantum computation, and show how some of its formal
properties lead to particular issues in for the rewriting machinery.

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects] the self dual object $Q$;
\item[Arrows] Two families of arrows:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : b
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down.  We use colours to denote two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  (The adjoints
for $H$, $\alpha_X$ and $\alpha_Z$ will be defined equationally.)  The
free compact closed category  is then given by all graphs formed by
composing and tensoring these basic graphs.

To model the behaviour of quantum systems certain additional equations
must be satisfied.  These are discussed in detail in
\cite{Coecke:2008jo};  we present them in graphical form in
Figure~\ref{fig:graph-quant-eqns}.

\begin{figure}
  \figureline

  The $(\cdot)^\dag$ functor gives a vertical symmetry to the
  category, hence for every equation we have a second equation
  obtained by flipping the diagram upside down.  In additon, we have a
  ``colour symmetry'' so for each equation shown here gives rise to
  second obtained by exchanging the two colours.
  \begin{description}
  \item[Comonoid Laws] 
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{3em}{comonoid-assoc1}  &=\;&
      \inlinegraphic{3em}{comonoid-assoc2}
      &\qquad\qquad&
      \inlinegraphic{3em}{comonoid-unit1}  &=\;&
      \inlinegraphic{3em}{comonoid-unit2}  &=\;&
      \inlinegraphic{3em}{comonoid-unit3} 
      &\qquad\qquad&
      \inlinegraphic{3em}{comonoid-comm1}  &=\;&
      \inlinegraphic{3em}{comonoid-comm2}
    \end{array}
    \]
  \item[Isometry, Frobenius, and  Compact Structure]
    \[
    \begin{array}{cccccccccccccccc}
      \inlinegraphic{3.5em}{isometry1}  &=\;&
      \inlinegraphic{3.5em}{isometry2}  
      &\qquad\qquad &
      \inlinegraphic{3em}{frobenius1}  &=\;&
      \inlinegraphic{3em}{frobenius2}  
      &\qquad\qquad&
      \inlinegraphic{2.5em}{compact1}  &=\;&
      \inlinegraphic{1.3em}{compact2}  
    \end{array}
    \]
  \item[Abelian Unitary Group]
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{2.2em}{group1}  &:=\;&
      \inlinegraphic{2.2em}{group2}  &=\;&
      \inlinegraphic{2.2em}{group3} 
      &\qquad\qquad&
      \left( \inlinegraphic{2.2em}{group4}\!\right)^\dag  &=\;&
      \inlinegraphic{2.2em}{group5} 
      &\qquad\qquad&
      \inlinegraphic{3.5em}{group6}  &=\;&
      \inlinegraphic{3.5em}{group7}  &=\;&
      \inlinegraphic{3.5em}{group8} 
    \end{array}
    \]
  \item[Bilinearity] 
    \[
    \begin{array}{cccccccccccc}
      \inlinegraphic{3.5em}{alpha-commute1}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute2}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute3} 
    \end{array}
    \]
  \item[Bialgebra Laws] Let  $\inlinegraphic{1em}{bialgebra1} := 
    \inlinegraphic{1.5em}{bialgebra2}$;  then:
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{3.5em}{bialgebra3}  &=\;&
      \inlinegraphic{3em}{bialgebra4}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra5}  &=\;&
      \inlinegraphic{1.5em}{bialgebra6}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra7}  &=\;&
      \inlinegraphic{1.5em}{bialgebra8} 
  \end{array}
  \]
\item[Group Actions]
  \[
  \begin{array}{cccccccccccccccccccccc}
    \inlinegraphic{2.5em}{group-int1} &=\;&
    \inlinegraphic{2.5em}{group-int2} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int3} &=\;&
    \inlinegraphic{2em}{group-int4} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int5} &=\;&
    \inlinegraphic{1.5em}{group-int6} 
    &\qquad\qquad&
    \inlinegraphic{2.5em}{group-int7} &=\;&
    \inlinegraphic{1.5em}{group-int8} 
  \end{array}
  \]
    \item[$H$ Properties]
    \[
    \begin{array}{cccccccccccccccccccc}
      \left(\inlinegraphic{1.5em}{H}\right)^\dag &=\;&
      \inlinegraphic{1.5em}{H} 
      &\qquad\qquad&
      \inlinegraphic{2.5em}{Heq1} &=\;&
      \inlinegraphic{2.5em}{Heq2}
    \end{array}
    \]
  \item[Colour Duality]
        \[
    \begin{array}{cccccccccccccccccccc}
      \inlinegraphic{2.5em}{Heq-delta} &=\;&
      \inlinegraphic{1.5em}{reddelta}
      &\qquad\qquad&
      \inlinegraphic{2em}{Heq-epsilon} &=\;&
      \inlinegraphic{1.5em}{redepsilon}
      &\qquad\qquad&
      \inlinegraphic{2.5em}{Heq-alpha} &=\;&
      \inlinegraphic{1.5em}{redalpha}
    \end{array}
    \]

  \end{description}

  \caption{Graphical Equations for Quantum Systems}
  \label{fig:graph-quant-eqns}
  \figureend
\end{figure}
